---
title: "Practice time!"
output:
  html_document:
    highlight: kate
    theme: lumen
    toc: no
    toc_float: no
  pdf_document:
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = T, cache = T, fig.width = 4, fig.height = 3, fig.align = "center")
```

# Hierarchical clustering

## Data

In this session, we are going to continue using the `protein.csv` data set from previous session.
We start by loading and preparing the data set.

```{r}
data.protein = read.csv("source/protein.csv")
rownames(data.protein) = data.protein$Country
data.protein$Country = NULL
data.protein = as.data.frame(scale(data.protein))
```

## Creating an agglomerative cluster

To create a hierarchical agglomerative cluster, we will use the function `hclust` from R base. The basic syntax of this function is `hclust(d, method)`, where `d` is a matrix of pairwise distaces between the observations and `method` describes the criterion used to join different clusters.

To compute `d`,  we can use the function `dist`.

```{r}
d = dist(data.protein)
```
```{r, echo=F}
knitr::kable(as.matrix(d)[1:6, 1:6])
```

Function `dist` by default computes the euclidean distance, but we can specify other distances as well, with the parameter `method`.

```{r}
set.seed(2018)
hc = hclust(d, method = "complete")
hc
```

We can plot the resultant dendrogram with the package `factoextra`.

```{r}
library(factoextra)

fviz_dend(hc, k = 4, cex=0.5)
```

### Linkages: Parameter `method` in `hclust`

Varying the parameter `method` in `hclust` function call, produces different dendrogams.

```{r}
hc = hclust(d, method = "single")
fviz_dend(hc, k = 4, cex=0.5)+labs(subtitle = "Single linkage")


hc = hclust(d, method = "average")
fviz_dend(hc, k = 4, cex=0.5)+labs(subtitle = "Average linkage")

```

### Method `cutree`

Function `cutree` makes a horizontal cut in the dendrogram. We can provide either one of `k` (number of groups) or `h` (height at which to cut the tree)

```{r}
cutree(hc, k = 4)
```

Notice that `cutree` returns a named vector with the same structure as field `cluster` returned by the function `kmeans`.

# Color quantization with kmeans

In this part, we will illustrate the use of kmeans in image segmentation (color quantization). We will use the `OpenImageR` library for R.

We start by downloading and reading the picture into R.

```{r}
library(OpenImageR)

#url = 'http://kiwi.uc3m.es/datos-curso-deloitte/Cave1.png'
#download.file(url, destfile = 'Cave1.png')

img = readImage('source/Cave1.png')
dim(img)
```

This picture has a resolution of `514 x 381` px, separated in three color chanels (R, G, B), thus we actually have `381*514=195834` observations and `3` variables. We can use the function `imageShow` to plot the picture.

```{r}
imageShow(img)
```

Since this is an example, we are going to resize this picture to a lower resolution, so `kmeans` runs faster.

```{r}
img1 = resizeImage(img, width = 190, height = 257, method = "bilinear")
dim(img1)
imageShow(img1)
```

Next, we transform our image, which is a `160 x 267 x 3` array, into a `42720 x 3` matrix.

```{r}
img1.vect = apply(img1, 3, as.vector)
dim(img1.vect)
```

We are now ready to cluster our data using `kmeans`.

```{r}
set.seed(2018)
km = kmeans(img1.vect, centers = 4)
```

Next, we substitute each pixel by the center of its cluster, and take the new segmented image to the original format.

```{r}
img2 = km$centers[km$cluster,]
dim(img2) = dim(img1)

imageShow(img2)
```

